---
output:
  html_document:
    keep_md: yes
---
# Reproducible Research: Peer Assessment 1


## Setting up the local working environment

Set the current working directory and if the data set is not already present, download it from the course link.

```{r}
my.path <- "C:/Coursera/Data Science - Specialization/Reproducible Research/RepData_PeerAssessment1";
setwd(my.path);

if (file.exists("activity.csv") == F) {
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip";
        fileZip <- "./activity.zip";
        if (file.exists(fileZip) == F) {
                download.file(fileUrl, fileZip, mode = "wb")
        }
}
```



## Loading and preprocessing the data

Accomodate for reading data from the extracted csv file or directly from the zip file.

```{r}
if (file.exists("activity.csv") == TRUE){
        df.org <- read.csv("activity.csv")
} else{
        df.org <- read.csv(unz(fileZip,"activity.csv"))
}

# Convert the date into POSIXct
df.org$date <- as.POSIXct(df.org$date)

# Convert the interval into factor
df.org$interval <- as.factor(df.org$interval)

str(df.org)
```

Using the reshape package's melt and cast functions, transform the original data set into required format to answer the questions

```{r}
require(reshape)

molten <- melt(df.org, id.vars=c("date", "interval"), na.rm=TRUE)
str(molten)
```

Now that we have the molten data from the original data set, we can calculate the aggregate per day
```{r}
df.total.steps.per.day <- cast(molten, date ~ variable, sum)
```


## What is mean total number of steps taken per day?

A histogram of the total number of steps per day.
```{r, Steps_Per_Day, fig.height=6}
par(mar = c(5,4,1,1), las=1)
hist(df.total.steps.per.day$steps, breaks=length(df.total.steps.per.day$date), main = "Total steps per day", xlab="Ticks")
```

We calculate the mean and median of the total number of steps tapen per day.
```{r}
mean.total.steps.per.day <- as.integer(mean(df.total.steps.per.day$steps))
median.total.steps.per.day <- as.integer(median(df.total.steps.per.day$steps))
```

```
Mean total number of steps per day = `r sprintf("%04d", mean.total.steps.per.day)`
Median number of steps per day = `r sprintf("%04d", median.total.steps.per.day)`
```

## What is the average daily activity pattern?

Fist we cast the molten data into a new data frame which can help us answer this question
```{r}
steps.per.interval <- cast(molten, interval ~ variable, mean)
```

We plot a time series of the 5-minute interval (x-axis) and the average number of steps taken, averaged accross all days (y-axis)
```{r, Mean_Steps_Per_Interval, fig.height=6}
require(lattice)

xaxis.ticks <- seq(0, nrow(steps.per.interval), by = 30)
xyplot(steps ~ interval, data = steps.per.interval, main = "Mean Steps Per Interval", layout=c(1,1), type=c("l"), scales=list(x=list(at=xaxis.ticks)))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
interval.with.maxsteps <- with(steps.per.interval, interval[which(steps == max(steps))])
```

```
Interval with maximum steps = `r interval.with.maxsteps`
```

## Imputing missing values

Create a new data frame with the missing values replaced by the mean steps per interval
```{r}
df.imp <- df.org
df.imp$steps <- ifelse(is.na(df.org$steps) == TRUE, steps.per.interval$steps[steps.per.interval$interval %in% df.org$interval], df.org$steps)
str(df.imp)
```

Melt the new data frame
```{r}
molten.imp <- melt(df.imp, id=c("date", "interval"))
```

Now that we have the molten data from the imputed data set, we can calculate the aggregate per day
```{r}
dfimp.total.steps.per.day <- cast(molten.imp, date ~ variable, sum)
```


A histogram of the total number of steps per day.
```{r, Steps_Per_Day_With_Imputation, fig.height=6}
par(mar = c(5,4,1,1), las=1)
hist(dfimp.total.steps.per.day$steps, breaks=length(dfimp.total.steps.per.day$date), main = "Total steps per day", xlab="Ticks")
```

We calculate the mean and median of the total number of steps tapen per day.
```{r}
imp.mean.total.steps.per.day <- as.integer(mean(dfimp.total.steps.per.day$steps))
imp.median.total.steps.per.day <- as.integer(median(dfimp.total.steps.per.day$steps))
```

With imputed data
```
Mean total number of steps per day = `r sprintf("%04d", imp.mean.total.steps.per.day)`
Median number of steps per day = `r sprintf("%04d", imp.median.total.steps.per.day)`
```

With Origianl data
```
Mean total number of steps per day = `r sprintf("%04d", mean.total.steps.per.day)`
Median number of steps per day = `r sprintf("%04d", median.total.steps.per.day)`
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new variable representing the weekday
```{r}
df.imp$day <- as.factor(ifelse(weekdays(df.imp$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
#df.imp$interval <- as.factor(df.imp$interval)
```

Melt this again to cater for the new variable added.
```{r}
molten.new <- melt(df.imp, id=c("day", "interval"), measure=c("steps"))
weekday.cast <- cast(molten.new, day+interval~variable, mean)
```

Plot the average steps per interval for the weekend days and week days.
```{r, Weekend_Vs_Weekday, fig.height=6}
#Create the Weekend vs Weekday plot
require(lattice)
xaxis.ticks <- seq(0, nrow(weekday.cast), by = 30)
xyplot(steps ~ interval | day, data = weekday.cast, layout=c(1,2), type=c("l"), scales=list(x=list(at=xaxis.ticks)))
```

